---
title: "ML_project"
author: "Ciprian C CHIRILA"
date: "August 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION (from the project assignement)

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

# PREPARE the R environment and load the libraries needed

```{r}
rm(list = ls()) # clean workspace
library(caret)
```

# LOAD the datasets

```{r}
ptrain <- read.csv("pml-training.csv")
ptest  <- read.csv("pml-testing.csv")
```


# SPLIT the training data in two smaller chunks

We choose to work with a smaller chunk of data to test our approach (this is faster). So, we separate the orginal train set into a train and testing 
sets.

```{r}
set.seed(132)
inTrain <- createDataPartition(y=ptrain$classe, p=0.7, list=F)
ptrain1 <- ptrain[ inTrain, ]
ptrain2 <- ptrain[-inTrain, ]
```

# CLEAN the data: remove almost linearly-dependent data, those that are not defined, and the variables that can't be used to predict

We identify the almost linear-dependent variables, the ones that are not defined, and the ones that can not be used for prediction (like for example,
the data and time of the measurements).

```{r}

# which of the variables have almost zero variance

nz <- nearZeroVar(ptrain1)
ptrain1 <- ptrain1[, -nz]
ptrain2 <- ptrain2[, -nz]

# variables mostly NAs

mostNA <- sapply(ptrain1, function(x) mean(is.na(x))) > 0.95
ptrain1 <- ptrain1[, mostNA==F]
ptrain2 <- ptrain2[, mostNA==F]

# remove variables that can't ne used to predict (the first 5)

ptrain1 <- ptrain1[, -(1:5)]
ptrain2 <- ptrain2[, -(1:5)]

```

## Modelling

Here, we employ the Random Forests (very good for nonlinear data). We use 3-fold cross-validation. As we will see, already this model is able to achieve
very high prediction accuracy, soe we won't have to use any other machine learning models.

```{r}

# use 3-fold CV to select best tuning parameters
fitCtrl <- trainControl(method="cv", number=3, verboseIter=F)

# fit model on ptrain1
fit <- train(classe ~ ., data=ptrain1, method="rf", trControl=fitCtrl)

# display the final model' details
fit$finalModel
```

```{r}
# use model to predict classe in validation set (ptrain2)
preds <- predict(fit, newdata=ptrain2)

# show confusion matrix to get estimate of out-of-sample error
confusionMatrix(ptrain2$classe, preds)
```

## Re-train the model using the original (full) sets of data

Since we saw that Random Forrest is a successfull choice, we do the same as above on the full sets provided (slower!).

```{r}

# take care of the variables with close to zero variance
nz <- nearZeroVar(ptrain)
ptrain <- ptrain[, -nz]
ptest <- ptest[, -nz]

# remove NAs
mostNA <- sapply(ptrain, function(x) mean(is.na(x))) > 0.95
ptrain <- ptrain[, mostNA==F]
ptest <- ptest[, mostNA==F]

# remove variables that can't be used to predict (first 5 vars, for example the time and date of measurement)
ptrain <- ptrain[, -(1:5)]
ptest <- ptest[, -(1:5)]

# refit model on the full training set (ptrain)
fitCtrl <- trainControl(method="cv", number=3, verboseIter=F)
fit <- train(classe ~ ., data=ptrain, method="rf", trControl=fitCtrl)

```

## Predict

```{r}

# predict on test set
predictions <- predict(fit, newdata=ptest)

# see the final predictions on the provided test set of 20 cases
results <- data.frame(problem_id=ptest$problem_id,predicted=predictions)
print(results)

```
